---
title: "Data Science Jobs Analysis"
subtitle: "Updated: `r format(Sys.Date(), '%B %d, %Y')`"
toc: true
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    page-layout: full
    grid:
      sidebar-width: 0px
      body-width: 1100px
      margin-width: 200px
      gutter-width: 1.5rem
execute:
  warning: false
---

## Data ETL and Report Pipeline

```{mermaid}
flowchart LR
  A[Web Scraped Data] --> B(Clean scraped data)
  B --> C(get domain of companies using chatGPT API service)
  C --> D(Normalize and Push Data into Postgres)
  D --> E[Generate Analysis in a Quarto HTML Report]
```

## Load Libraries 

```{r}
library(ggplot2)
library(gt)
library(gtExtras)
library(maps)
library(tidygeocoder)
library(tidyverse)
library(readr)
library(stringr)
library(jsonlite)
library(dbplyr)
library(DBI)
```

## Pull Data from Postgres

```{r}
pw <- "muix7pcj"

con <- dbConnect(
  RPostgres::Postgres(), 
  dbname = "ds_jobs",
  port = 5432, 
  user = "postgres", 
  password = pw
  )
```

## Dataset overview

### First 5 rows
```{r}

knitr::kable(head(dbGetQuery(con, "SELECT * FROM jobpostings;"),5))

knitr::kable(head(dbGetQuery(con, "SELECT * FROM skills_table;"),5))

knitr::kable(head(dbGetQuery(con, "SELECT * FROM skill_categories;"),5))

knitr::kable(head(dbGetQuery(con, "SELECT * FROM job_skill_mapping;"),5))
```

### Column type Main Table
```{r}
str(dbGetQuery(con, "SELECT * FROM jobpostings;"))
```
### Number of Rows and Cols

```{r}
nrow(dbGetQuery(con, "SELECT * FROM jobpostings;"))

nrow(dbGetQuery(con, "SELECT * FROM skills_table;"))

nrow(dbGetQuery(con, "SELECT * FROM skill_categories;"))

nrow(dbGetQuery(con, "SELECT * FROM job_skill_mapping;"))
```

# 5 Questions answered

Each question is answered by querying the needed data from the Postgres database. This will involve joining multiple tables to get the desired data.

## Q1. What are the main technologies? 

```{r}
#| code-fold: true

# skill_vector <- unlist(strsplit(gsub("\\[|\\]|'", "", df$job_skills), ",\\s*"))
# freq_table <- as.data.frame(table(skill_vector))
# colnames(freq_table) <- c("skill", "n")
# 
# freq_table <- freq_table[order(-freq_table$n), ]
# top_20 <- head(freq_table, 20)
# top_20$skill <- factor(top_20$skill, levels = top_20$skill[order(top_20$n, decreasing = FALSE)])

plot_df <- dbGetQuery(con, "WITH SkillCount AS (
    SELECT st.skill,
    COUNT(jsm.jobid) AS job_count
    FROM job_skill_mapping jsm
    JOIN skills_table st ON jsm.skill_id = st.skill_id
    GROUP BY st.skill
)
SELECT skill, job_count
FROM SkillCount
ORDER BY job_count DESC
LIMIT 20;")

# most popular technologies
ggplot(plot_df, aes(y = reorder(skill, job_count), x = job_count)) +
  geom_bar(stat = "identity", fill = "firebrick3") +
  scale_x_continuous(breaks = c(0,2500,5000,7500,10000,12500)) +
  theme_minimal() +
  labs(title = "Top 20 Skills",
       x = "Frequency",
       y = "Skill")

ggsave("../output/top20_skills.png", dpi = 400, bg = "#FFFFFF")
```

### What are the main technologies by category?

```{r}
#| code-fold: true

color_vec <- c("firebrick3", "#00A08A", "#F98400", "#5BBCD6", "#C6CDF7", "#000000", "#CDC08C", "#004990", "grey", "darkseagreen4")

# most popular technologies by category
# v_cleaned <- gsub("'", "\"", df$job_type_skills)
# 
# parsed_list <- lapply(v_cleaned, fromJSON)
# 
# keys <- c()
# values <- c()
# 
# for (item in parsed_list) {
#   for (key in names(item)) {
#     keys <- c(keys, key)  # Append the key
#     values <- c(values, paste(item[[key]], collapse = ", "))
#   }
# }
# 
# key_value_df <- data.frame(Key = keys, Values = values)
# 
# separate_skill_type <- function(i, key_value_df){
#   skill_vector <- unlist(strsplit(gsub("\\[|\\]|'", "", key_value_df[key_value_df$Key == i,]$Values), ",\\s*"))
#   freq_table <- as.data.frame(table(skill_vector))
#   colnames(freq_table) <- c("skill", "n")
#   freq_table$type <- i
#   
#   return(freq_table)
# }
# 
# skill_type_freq <- data.frame(skill = character(), n = integer(), type = character())
# 
# for(i in unique(key_value_df$Key)){
#   skill_type_freq <- rbind(skill_type_freq, separate_skill_type(i, key_value_df))
# }
# 
# df_top_10 <- skill_type_freq %>%
#   group_by(type) %>%
#   arrange(desc(n)) %>%
#   slice_head(n = 10) %>%
#   ungroup()

plot_df <- dbGetQuery(con, "WITH SkillCount AS (
SELECT sc.skill_category, st.skill,
COUNT(jsm.jobid) AS job_count
FROM job_skill_mapping jsm
JOIN skills_table st ON jsm.skill_id = st.skill_id
JOIN skill_categories sc ON st.skill_id = sc.skill_id
GROUP BY sc.skill_category, st.skill
), 
RankedSkills AS (SELECT skill_category, skill, job_count,
RANK() OVER (PARTITION BY skill_category ORDER BY job_count DESC) AS rank
FROM SkillCount
)
SELECT skill_category, skill, job_count
FROM RankedSkills
WHERE rank <= 10
ORDER BY skill_category, rank;")

g <- ggplot(plot_df, aes(x = reorder(skill, job_count), y = job_count, fill = skill_category)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500), limits = c(0, 13600), expand = c(0, 0)) +
  geom_text(aes(label = job_count,  colour = skill_category), vjust = 0.3, hjust = -0.3, size = 2.5, show.legend = FALSE) +
  facet_wrap(~ skill_category, scales = "free_y", ncol = 2) +
  coord_flip() + 
  scale_fill_manual(values = color_vec) +
  scale_colour_manual(values = color_vec) +
  theme_minimal() +
  labs(x = "Skill", y = "Count", title = "Top 10 Skills by Skill Category") +
  guides(fill = "none") 
```


```{r}
#| fig-width: 8
#| fig-height: 12
#| fig-format: svg

g

ggsave("../output/top10_skill_cat.png", dpi = 400, bg = "#FFFFFF")
```


## Q2. What is the median salary for each job title?  

```{r}
#| code-fold: true

plot_df <- dbGetQuery(con, "SELECT job_title_short, salary_year_avg FROM jobpostings;")

plot_df <- plot_df %>% 
      group_by(job_title_short) %>%
      summarise(med_salary = round(median(salary_year_avg),0)) %>%
      arrange(desc(med_salary))

# avg salary for each role
ggplot(plot_df, aes(y = med_salary, x = reorder(job_title_short, med_salary))) +
  geom_bar(stat = "identity", fill = "firebrick3") +
  scale_y_continuous(limits = c(0, 155000), labels = scales::label_dollar(scale_cut = scales::cut_short_scale()), expand = c(0,0)) +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Median Salary for Each Role",
       y = "Median Salary (USD)",
       x = "Job Title")

ggsave("../output/med_salary_role.png", dpi = 400, bg = "#FFFFFF")
```

```{r}
#| code-fold: true

# complicated string parsing avoided cause we normalized our data

# # Parse the job_type_skills vector and extract the relevant data
# v_cleaned <- gsub("'", "\"", df$job_type_skills)
# parsed_skills = lapply(v_cleaned, function(x) fromJSON(x))
# 
# # Initialize vectors to store the final data
# skill = c()
# salary = c()
# type = c()
# salary_year_avg = df$salary_year_avg
# 
# # Loop through each parsed skill and its corresponding salary
# for(i in 1:length(parsed_skills)){
#   # Get the skill types and corresponding skills
#   job_types = names(parsed_skills[[i]])
#   skills = unlist(parsed_skills[[i]])
#   
#   # Append the data to the final vectors
#   skill = c(skill, skills)
#   salary = c(salary, rep(salary_year_avg[i], length(skills)))
#   type = c(type, rep(job_types, sapply(parsed_skills[[i]], length)))
# }
# 
# # Combine into a data frame
# plot_df = cbind.data.frame(skill = skill, salary = salary, type = type)
# plot_df[plot_df$skill == "no-sql",]$skill <- "nosql"


plot_df <- dbGetQuery(con, 
"SELECT sc.skill_category, st.skill, jp.salary_year_avg
FROM job_skill_mapping jsm
JOIN skills_table st ON jsm.skill_id = st.skill_id
JOIN skill_categories sc ON st.skill_id = sc.skill_id
JOIN jobpostings jp ON jsm.jobid = jp.jobid;")

plot_df <- plot_df %>%
  group_by(skill_category, skill) %>%
  summarise(med_salary = round(median(salary_year_avg), 0), n = n(), 
            max_salary = round(max(salary_year_avg),0), min_salary = round(min(salary_year_avg),0))

plot_df <- plot_df[plot_df$n >= 50,]

plot_df <- plot_df %>%
  group_by(skill_category) %>%
  arrange(desc(med_salary)) %>%
  slice_head(n = 20) %>%
  ungroup()

g <- ggplot(plot_df, aes(x = reorder(skill, med_salary), y = med_salary, fill = skill_category)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(expand = c(0, 0), labels = scales::label_dollar(scale_cut = scales::cut_short_scale())) +
  geom_text(aes(label = med_salary,  colour = skill_category), 
            vjust = 0.3, hjust = -0.3, size = 2.5, show.legend = FALSE) +
  facet_wrap(~ skill_category, scales = "free_y", ncol = 2) +
  coord_flip(ylim = c(70000,185000)) + 
  scale_fill_manual(values = color_vec) +
  scale_colour_manual(values = color_vec) +
  theme_minimal() +
  labs(x = "Skill", y = "Median Salary", title = "Top 20 Skills by Skill Category") +
  guides(fill = "none") + 
  theme(panel.grid.minor.x = element_blank())
```


```{r}
#| fig-width: 10
#| fig-height: 14
#| fig-format: svg

g

ggsave("../output/top20skill_salary_cat.png", dpi = 400, bg = "#FFFFFF")
```


## Q3. Which industry field is asking for these roles ? 

```{r}
#| code-fold: true
#| label: tbl-industry
#| tbl-cap: "Company Domain"

plot_df <- dbGetQuery(con, "SELECT industry FROM jobpostings;")

DT::datatable(
  plot_df %>%
    filter(!is.na(industry)) %>%
    group_by(industry) %>%
    summarise(Count = n()) %>%
    #top_n(50, wt=Count) %>%
    arrange(desc(Count)),
  options = list(
    pageLength = 15
  )
  )
```

## Q4. What is the relation of location with respect to the number of job postings and salary range?

Salary varies by job country location, and job title. We will focus on United States jobs and specifically on data scientist role.

```{r}
#| code-fold: true

plot_df <- dbGetQuery(con, "SELECT job_location, salary_year_avg FROM jobpostings
                      WHERE job_country = 'United States' AND job_title_short = 'Data Scientist';")

#plot_df <- df[df$job_country == 'United States' & df$job_title_short == 'Data Scientist',]
plot_df$job_location <- sub(",.*", "", plot_df$job_location)

plot_df <- plot_df %>%
    group_by(job_location) %>%
    summarise(n = n(), med_salary = median(salary_year_avg)) %>%
    filter(n >= 10) %>%
    arrange(desc(med_salary))

`%ni%` <- Negate(`%in%`)

plot_df <- plot_df[plot_df$job_location %ni% c('Toronto', 'United States', 'Anywhere', 'California'),]
plot_df$job_location <- paste(plot_df$job_location, ', United States', sep = '')

# Geocode the cities
locations <- tibble(city = plot_df$job_location) %>% geocode(city, method = "osm", lat = latitude, long = longitude)

colnames(plot_df) <- c('city', 'n', 'med_salary')
plot_df <- merge(plot_df, locations, by='city')

write.csv(plot_df, '../data/us_cities_lat_long.csv', row.names = FALSE)
#plot_df <- read.csv('../data/us_cities_lat_long.csv')

plot_df$city <- sub(",.*", "", plot_df$city)

# Get world map data
usa_map <- map_data("state")
library(ggrepel)

plot_df %>%
  arrange(desc(med_salary))

g <- ggplot() +
  # Plot USA map
  geom_polygon(data = usa_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  # Overlay city points
  geom_point(data = plot_df, aes(x = longitude, y = latitude, size = n, color = med_salary), alpha = 0.8) +
  # Add city labels with ggrepel
  geom_text_repel(data = plot_df, aes(x = longitude, y = latitude, label = city), 
                size = 2.5, nudge_y = 1, segment.color = "gray50", segment.size = 0.5, max.overlaps=20) +
  # Customize appearance
  scale_color_gradient(low = "#004990", high = "firebrick2", name = "Median Salary") +
  scale_size_continuous(range = c(3, 10), name = "# of Positings") +
  theme_minimal(base_size = 14) +
  labs(title = "Average Salary of a Data Scientist by City", x = "Longitude", y = "Latitude") +
  coord_fixed(1.3) # Fix aspect ratio
```



```{r}
#| fig-width: 12
#| fig-height: 8

g

ggsave("../output/map.png", dpi = 400, bg = "#FFFFFF")
```


## Q5. Does a remote / work from home contract give on average lesser pay than an onsite contract?

The remote jobs are US origin jobs (not exactly labelled if you can work anywhere in the world or anywhere in the US), so we will for now compare it with onsite US jobs.

We will use Full-time contract to avoid large amount of variance between internship, short contracts vs full-time contracts.

```{r}
#| code-fold: true

online <- dbGetQuery(con, "SELECT job_title_short, job_location, search_location, salary_year_avg FROM jobpostings
                     WHERE job_work_from_home = FALSE AND job_country = 'United States' AND job_schedule_type = 'Full-time';")

#online <- df[df$job_work_from_home == FALSE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
online$job_location <- sub(".*,\\s*", "", online$job_location)
online$job_location <- gsub("\\s*\\(.*?\\)", "", online$job_location)

online[online$job_location == "NY",]$job_location <- "New York"
online[online$job_location == "GA",]$job_location <- "Georgia"
online[online$job_location == "FL",]$job_location <- "Florida"
online[online$job_location == "IL",]$job_location <- "Illinois"
online[online$job_location == "CA",]$job_location <- "California"
online[online$job_location == "TX",]$job_location <- "Texas"

online <- online %>%
        group_by(job_location, job_title_short) %>%
        summarise(n=n(), med_salary = median(salary_year_avg)) %>%
        arrange(job_location)

offline <- dbGetQuery(con, "SELECT job_title_short, job_location, search_location, salary_year_avg FROM jobpostings
                      WHERE job_work_from_home = TRUE AND job_country = 'United States' AND job_schedule_type = 'Full-time';")

#offline <- df[df$job_work_from_home == TRUE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
offline$search_location <- sub(",.*", "", offline$search_location)
offline <- offline %>%
  group_by(search_location, job_title_short) %>%
  summarise(n=n(), med_salary = median(salary_year_avg)) %>%
  arrange(search_location)

states <- c("California","Florida","Georgia","Illinois","New York","Texas")
jobs <- c("Business Analyst","Data Analyst","Data Engineer","Data Scientist","Machine Learning Engineer",
          "Senior Data Analyst","Senior Data Engineer","Senior Data Scientist","Software Engineer")

diff_df <- as.data.frame(jobs)

for(i in states){
  temp_df <- merge(offline[offline$search_location == i,], online[online$job_location == i,],by="job_title_short") 
  temp_df <- temp_df %>% group_by(job_title_short) %>% 
    summarise(diff_offline_online = round((med_salary.x - med_salary.y)/med_salary.y,2))
  colnames(temp_df) <- c('job', paste(i,"_off_on", sep = ""))
  diff_df <- cbind(diff_df, temp_df[,2])
}

# online <- df[df$job_work_from_home == FALSE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
# online <- online %>%
#   group_by(job_title_short) %>%
#   summarise(n=n(), med_salary = median(salary_year_avg)) %>%
#   arrange(job_title_short)
# 
# offline <- df[df$job_work_from_home == TRUE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
# offline <- offline %>%
#       group_by(job_title_short) %>%
#       summarise(n=n(), med_salary = median(salary_year_avg)) %>%
#       arrange(job_title_short)
# 
# diff_df <- merge(offline, online,by="job_title_short")
# diff_df <- diff_df %>% group_by(job_title_short) %>% 
#       summarise(diff_offline_online = round(((med_salary.x - med_salary.y)/med_salary.y)*100,2))
# 
# diff_df
```

```{r}
#| label: tbl-remote
#| tbl-cap: "Remote vs On-Site Salary Difference"

gt(diff_df) %>%
  cols_label(
    jobs = "Job Title",
    California_off_on = "California",
    Florida_off_on = "Florida",
    Georgia_off_on = "Georgia",
    Illinois_off_on = "Illinois",
    `New York_off_on` = "New York",
    Texas_off_on = "Texas"
  ) %>%
  tab_header(
    title = "Remote vs On-Site Salary Difference"
  ) %>% 
  fmt_percent(
    columns = California_off_on:Texas_off_on,
    decimals = 2
  ) %>%
  data_color(
    columns = California_off_on:Texas_off_on,
    fn = \(v) lapply(v, \(x) if (is.na(x)) "grey" else if (x < 0) "darkred" else if (x > 0) "darkgreen" else "grey"),
    apply_to = "text"
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")), # Bold text
    locations = cells_title(groups = "title")
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")), # Bold text
    locations = cells_column_labels()
  )

dbDisconnect(con)
```
