---
title: "Data Science Jobs Analysis"
subtitle: "Updated: `r format(Sys.Date(), '%B %d, %Y')`"
toc: true
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    page-layout: full
    grid:
      sidebar-width: 0px
      body-width: 1100px
      margin-width: 200px
      gutter-width: 1.5rem
execute:
  warning: false
---

## Data ETL and Report Pipeline

```{mermaid}
flowchart LR
  A[Web Scraped Data in csv file] --> B(Clean scraped data)
  B --> C(get domain of company using chatGPT API service)
  C --> D(Merge and Push Data into Postgres)
  D --> E[Pull Data from Postgres and Render Analysis report using Quarto]
```

## Load Libraries 

```{r}
library(ggplot2)
library(gt)
library(gtExtras)
library(maps)
library(tidygeocoder)
library(tidyverse)
library(readr)
library(stringr)
library(jsonlite)
library(dbplyr)
library(DBI)
library(ggrepel)
```

## Pull Data from Postgres

```{r}

pw <- "muix7pcj"

con <- dbConnect(
  RPostgres::Postgres(), 
  dbname = "ds_jobs",
  port = 5432, 
  user = "postgres", 
  password = pw
  )

df <- dbGetQuery(con, "SELECT * FROM jobs;")
dbDisconnect(con)

write.csv(df, '../data/pulled_data.csv', row.names = FALSE)
```

## Dataset overview

### First 5 rows
```{r}
knitr::kable(head(df,5))
```

### Column types
```{r}
str(df)
```

## Q1. What are the main technologies? 

```{r}
#| code-fold: true

skill_vector <- unlist(strsplit(gsub("\\[|\\]|'", "", df$job_skills), ",\\s*"))
freq_table <- as.data.frame(table(skill_vector))
colnames(freq_table) <- c("skill", "n")

freq_table <- freq_table[order(-freq_table$n), ]
top_20 <- head(freq_table, 20)
top_20$skill <- factor(top_20$skill, levels = top_20$skill[order(top_20$n, decreasing = FALSE)])

write.csv(top_20, '../data/top20_skills.csv', row.names = FALSE)

# most popular technologies
ggplot(top_20, aes(y = skill, x = n)) +
  geom_bar(stat = "identity", fill = "firebrick3") +
  scale_x_continuous(breaks = c(0,2500,5000,7500,10000,12500)) +
  theme_minimal() +
  labs(title = "Top 20 Skills",
       x = "Frequency",
       y = "Skill")
```

### What are the main technologies by category?

```{r}
#| code-fold: true

# most popular technologies by category
v_cleaned <- gsub("'", "\"", df$job_type_skills)

parsed_list <- lapply(v_cleaned, fromJSON)

keys <- c()
values <- c()

for (item in parsed_list) {
  for (key in names(item)) {
    keys <- c(keys, key)  # Append the key
    values <- c(values, paste(item[[key]], collapse = ", "))
  }
}

key_value_df <- data.frame(Key = keys, Values = values)

separate_skill_type <- function(i, key_value_df){
  skill_vector <- unlist(strsplit(gsub("\\[|\\]|'", "", key_value_df[key_value_df$Key == i,]$Values), ",\\s*"))
  freq_table <- as.data.frame(table(skill_vector))
  colnames(freq_table) <- c("skill", "n")
  freq_table$type <- i
  
  return(freq_table)
}

skill_type_freq <- data.frame(skill = character(), n = integer(), type = character())

for(i in unique(key_value_df$Key)){
  skill_type_freq <- rbind(skill_type_freq, separate_skill_type(i, key_value_df))
}

df_top_10 <- skill_type_freq %>%
  group_by(type) %>%
  arrange(desc(n)) %>%
  slice_head(n = 10) %>%
  ungroup()

color_vec <- c("firebrick3", "#00A08A", "#F98400", "#5BBCD6", "#C6CDF7", "#000000", "#CDC08C", "#004990", "grey", "darkseagreen4")

write.csv(df_top_10, '../data/top10_skills_cat.csv', row.names = FALSE)

g <- ggplot(df_top_10, aes(x = reorder(skill, n), y = n, fill = type)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500), limits = c(0, 13600), expand = c(0, 0)) +
  geom_text(aes(label = n,  colour = type), vjust = 0.3, hjust = -0.3, size = 2.5, show.legend = FALSE) +
  facet_wrap(~ type, scales = "free_y", ncol = 2) +
  coord_flip() + 
  scale_fill_manual(values = color_vec) +
  scale_colour_manual(values = color_vec) +
  theme_minimal() +
  labs(x = "Skill", y = "Count", title = "Top 10 Skills by Skill Category") +
  guides(fill = "none") 
```


```{r}
#| fig-width: 8
#| fig-height: 12
#| fig-format: svg

g
```


## Q2. What is the median salary for each job title?  

```{r}
#| code-fold: true

plot_df <- df %>% 
      group_by(job_title_short) %>%
      summarise(med_salary = round(median(salary_year_avg),0)) %>%
      arrange(desc(med_salary))

write.csv(plot_df, '../data/median_salary_role.csv', row.names = FALSE)

# avg salary for each role
ggplot(plot_df, aes(y = med_salary, x = reorder(job_title_short, med_salary))) +
  geom_bar(stat = "identity", fill = "firebrick3") +
  scale_y_continuous(limits = c(0, 155000), labels = scales::label_dollar(scale_cut = scales::cut_short_scale()), expand = c(0,0)) +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Median Salary for Each Role",
       y = "Median Salary (USD)",
       x = "Job Title")

```

```{r}
#| code-fold: true

# which technical skill gives better pay?
# Are there any niche, less popular tech skills that have high pay?

# Parse the job_type_skills vector and extract the relevant data
v_cleaned <- gsub("'", "\"", df$job_type_skills)
parsed_skills = lapply(v_cleaned, function(x) fromJSON(x))

# Initialize vectors to store the final data
skill = c()
salary = c()
type = c()
salary_year_avg = df$salary_year_avg

# Loop through each parsed skill and its corresponding salary
for(i in 1:length(parsed_skills)){
  # Get the skill types and corresponding skills
  job_types = names(parsed_skills[[i]])
  skills = unlist(parsed_skills[[i]])
  
  # Append the data to the final vectors
  skill = c(skill, skills)
  salary = c(salary, rep(salary_year_avg[i], length(skills)))
  type = c(type, rep(job_types, sapply(parsed_skills[[i]], length)))
}

# Combine into a data frame
plot_df = cbind.data.frame(skill = skill, salary = salary, type = type)
plot_df[plot_df$skill == "no-sql",]$skill <- "nosql"

plot_df_f <- plot_df %>%
    group_by(type, skill) %>%
    summarise(med_salary = round(mean(salary), 0), n = n(), max_salary = round(max(salary),0), min_salary = round(min(salary),0))

plot_df_f <- plot_df_f[plot_df_f$n >= 50,]

plot_df_f <- plot_df_f %>%
  group_by(type) %>%
  arrange(desc(med_salary)) %>%
  slice_head(n = 20) %>%
  ungroup()

write.csv(plot_df_f, '../data/median_salary_role_cat.csv', row.names = FALSE)

g <- ggplot(plot_df_f, aes(x = reorder(skill, med_salary), y = med_salary, fill = type)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(expand = c(0, 0), labels = scales::label_dollar(scale_cut = scales::cut_short_scale())) +
  geom_text(aes(label = med_salary,  colour = type), 
            vjust = 0.3, hjust = -0.3, size = 2.5, show.legend = FALSE) +
  facet_wrap(~ type, scales = "free_y", ncol = 2) +
  coord_flip(ylim = c(70000,185000)) + 
  scale_fill_manual(values = color_vec) +
  scale_colour_manual(values = color_vec) +
  theme_minimal() +
  labs(x = "Skill", y = "Median Salary", title = "Top 20 Skills by Type") +
  guides(fill = "none") + 
  theme(panel.grid.minor.x = element_blank())
```


```{r}
#| fig-width: 10
#| fig-height: 14
#| fig-format: svg

g
```


## Q3. Which industry field is asking for these roles ? 

```{r}
#| code-fold: true
#| label: tbl-industry
#| tbl-cap: "Company Domain"


DT::datatable(
  df %>%
    group_by(industry) %>%
    summarise(Count = n()) %>%
    filter(!is.na(industry)) %>%
    #top_n(50, wt=Count) %>%
    arrange(desc(Count)),
  options = list(
    pageLength = 15
  )
  )
```

## Q4. What is the relation of location with respect to the number of job postings and salary range?

```{r}
#| code-fold: true

plot_df <- df[df$job_country == 'United States' & df$job_title_short == 'Data Scientist',]
plot_df$job_location <- sub(",.*", "", plot_df$job_location)

plot_df <- plot_df %>%
    group_by(job_location) %>%
    summarise(n = n(), med_salary = median(salary_year_avg)) %>%
    filter(n >= 10) %>%
    arrange(desc(med_salary))

`%ni%` <- Negate(`%in%`)

plot_df <- plot_df[plot_df$job_location %ni% c('Toronto', 'United States', 'Anywhere', 'California'),]
plot_df$job_location <- paste(plot_df$job_location, ', United States', sep = '')

# Geocode the cities
locations <- tibble(city = plot_df$job_location) %>% geocode(city, method = "osm", lat = latitude, long = longitude)

colnames(plot_df) <- c('city', 'n', 'med_salary')
plot_df <- merge(plot_df, locations, by='city')

#plot_df <- read.csv('../data/us_cities_lat_long.csv')

plot_df$city <- sub(",.*", "", plot_df$city)

# Get world map data
usa_map <- map_data("state")

write.csv(plot_df, '../data/us_cities_lat_long.csv', row.names = FALSE)

g <- ggplot() +
  # Plot USA map
  geom_polygon(data = usa_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  # Overlay city points
  geom_point(data = plot_df, aes(x = longitude, y = latitude, size = n, color = med_salary), alpha = 0.8) +
  # Add city labels with ggrepel
  geom_text_repel(data = plot_df, aes(x = longitude, y = latitude, label = city), 
                size = 2.5, nudge_y = 1, segment.color = "gray50", segment.size = 0.5, max.overlaps=20) +
  # Customize appearance
  scale_color_gradient(low = "#004990", high = "firebrick2", name = "Median Salary") +
  scale_size_continuous(range = c(3, 10), name = "# of Positings") +
  theme_minimal(base_size = 14) +
  labs(title = "Average Salary of a Data Scientist by City", x = "Longitude", y = "Latitude") +
  coord_fixed(1.3) # Fix aspect ratio
```



```{r}
#| fig-width: 12
#| fig-height: 8

g
```


## Q5. Does a remote / work from home contract give on average lesser pay than an onsite contract?

```{r}
#| code-fold: true

online <- df[df$job_work_from_home == FALSE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
online$job_location <- sub(".*,\\s*", "", online$job_location)
online$job_location <- gsub("\\s*\\(.*?\\)", "", online$job_location)

online[online$job_location == "NY",]$job_location <- "New York"
online[online$job_location == "GA",]$job_location <- "Georgia"
online[online$job_location == "FL",]$job_location <- "Florida"
online[online$job_location == "IL",]$job_location <- "Illinois"
online[online$job_location == "CA",]$job_location <- "California"
online[online$job_location == "TX",]$job_location <- "Texas"

online <- online %>%
        group_by(job_location, job_title_short) %>%
        summarise(n=n(), med_salary = median(salary_year_avg)) %>%
        arrange(job_location)

offline <- df[df$job_work_from_home == TRUE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
offline$search_location <- sub(",.*", "", offline$search_location)
offline <- offline %>%
  group_by(search_location, job_title_short) %>%
  summarise(n=n(), med_salary = median(salary_year_avg)) %>%
  arrange(search_location)

states <- c("California","Florida","Georgia","Illinois","New York","Texas")
jobs <- c("Business Analyst","Data Analyst","Data Engineer","Data Scientist","Machine Learning Engineer",
          "Senior Data Analyst","Senior Data Engineer","Senior Data Scientist","Software Engineer")

diff_df <- as.data.frame(jobs)

for(i in states){
  temp_df <- merge(offline[offline$search_location == i,], online[online$job_location == i,],by="job_title_short") 
  temp_df <- temp_df %>% group_by(job_title_short) %>% 
    summarise(diff_offline_online = round((med_salary.x - med_salary.y)/med_salary.y,2))
  colnames(temp_df) <- c('job', paste(i,"_off_on", sep = ""))
  diff_df <- cbind(diff_df, temp_df[,2])
}

write.csv(diff_df, '../data/remote.csv', row.names = FALSE)

# online <- df[df$job_work_from_home == FALSE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
# online <- online %>%
#   group_by(job_title_short) %>%
#   summarise(n=n(), med_salary = median(salary_year_avg)) %>%
#   arrange(job_title_short)
# 
# offline <- df[df$job_work_from_home == TRUE & df$job_country == 'United States' & df$job_schedule_type == 'Full-time',]
# offline <- offline %>%
#       group_by(job_title_short) %>%
#       summarise(n=n(), med_salary = median(salary_year_avg)) %>%
#       arrange(job_title_short)
# 
# diff_df <- merge(offline, online,by="job_title_short")
# diff_df <- diff_df %>% group_by(job_title_short) %>% 
#       summarise(diff_offline_online = round(((med_salary.x - med_salary.y)/med_salary.y)*100,2))
# 
# diff_df
```

```{r}
#| label: tbl-remote
#| tbl-cap: "Remote vs On-Site Salary Difference"

gt(diff_df) %>%
  cols_label(
    jobs = "Job Title",
    California_off_on = "California",
    Florida_off_on = "Florida",
    Georgia_off_on = "Georgia",
    Illinois_off_on = "Illinois",
    `New York_off_on` = "New York",
    Texas_off_on = "Texas"
  ) %>%
  tab_header(
    title = "Remote vs On-Site Salary Difference"
  ) %>% 
  fmt_percent(
    columns = California_off_on:Texas_off_on,
    decimals = 2
  ) %>%
  data_color(
    columns = California_off_on:Texas_off_on,
    fn = \(v) lapply(v, \(x) if (is.na(x)) "grey" else if (x < 0) "darkred" else if (x > 0) "darkgreen" else "grey"),
    apply_to = "text"
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")), # Bold text
    locations = cells_title(groups = "title")
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")), # Bold text
    locations = cells_column_labels()
  )
```
